---
title: "SimBac Function"
author: "Haochu"
format: html
editor: visual
---

```{r}
#| label: load-packages
#| include: false

library(ggplot2)
library(microbenchmark)
library(patchwork)
library(sdprisk)
source("sim_gene/FSM/sim_FSM_ARG.r")
source("sim_gene/FSM/simbac_ARG.r")
source("sim_gene/FSM/sim_FSM_mutation.r")
source("sim_gene/ref/simu.R")
source("sim_gene/sim_birth_death.r")
source("sim_gene/local_tree.r")
source("sim_gene/ARG_igraph.r")
```

## Simulation Using `SimBac` Algorithm

### ARG Height

Compare heights of simulated ARGs from `simbac_ARG`, `sim_FSM_ARG(optimise=T)`, `sim_FSM_ARG(optimise=F)`. The red box represents the hitting time from a birth-death process.

```{r}
#| echo: true
time_df <- data.frame(time=rep(NA, 4000),
                      func=c(rep("Birth-death Process", 1000),
                             rep("sim_FSM_ARG(optimise=T)", 1000),
                             rep("sim_FSM_ARG(optimise=F)", 1000),
                             rep("simbac", 1000)))

set.seed(10)
for (i in 1:1000) {
  t <- sim_birth_death(100, 5)
  time_df$time[i] <- t
  
  r <- sim_FSM_ARG(100, 5, 100, bacteria = TRUE, delta = 5,
                   node_max = 100000, optimise_recomb = TRUE)
  time_df$time[1000+i] <- r$sum_time
  
  r <- sim_FSM_ARG(100, 5, 100, bacteria = TRUE, delta = 5,
                   node_max = 100000, optimise_recomb = FALSE)
  time_df$time[2000+i] <- r$sum_time
  
  r <- simbac_ARG(100, 5, 100, delta = 5, node_max = 100000, output_eff_R = TRUE)
  time_df$time[3000+i] <- r$sum_time
  
  if (i%%100 == 0) {print(paste("Complete", i, "iterations"))}
}
```

```{r}
#| echo: true
ggplot(time_df, aes(x=func, y=time, fill=func)) +
  geom_violin(width=1, alpha=0.5) +
  geom_boxplot(width=0.1, color="black", alpha=0.2) +
  labs(title = "# of leaf lineages = 100, rho = 5",
       x = "Functions",
       y = "ARG height") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 16),
    axis.text.x = element_text(size = 12),
    axis.title.y = element_text(size = 16),
    axis.text.y = element_text(size = 12),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12)
  )
```

### `simbac_ARG` Local Tree

Simulate full ARG by `simbac_ARG` and examine its height of local trees with the hypo-exponential distribution.

```{r}
#| echo: true
height_t_df <- data.frame(s1=rep(NA, 1000),
                          s50=rep(NA, 1000),
                          s80=rep(NA, 1000))
set.seed(100)
for (i in 1:1000) {
  ARG <- simbac_ARG(100, 5, 100, delta = 5, node_max = 100000, output_eff_R = TRUE)
  tree1 <- local_tree(ARG, 1)
  tree50 <- local_tree(ARG, 50)
  tree80 <- local_tree(ARG, 80)
  height_t_df$s1[i] <- tree1$sum_time
  height_t_df$s50[i] <- tree50$sum_time
  height_t_df$s80[i] <- tree80$sum_time
  if (i%%100 == 0) {print(paste("Complete", i, "iterations"))}
}
```

```{r}
#| echo: true
n <- 100
height_rate <- n:2 * (n-1):1 / 2
height_density <- function(x) {
  dhypoexp(x, rate=height_rate)
}

hist1 <- ggplot(height_t_df, aes(x = s1)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 fill = "lightgreen",
                 color = "black",
                 alpha = 0.7) +
  stat_function(fun = height_density, # Use your custom function here
                color = "darkblue",
                linewidth = 1.2,
                linetype = "solid") +
  labs(title = "Locat tree at site 1",
       x = "Height",
       y = "Density") +
  theme_minimal()
hist2 <- ggplot(height_t_df, aes(x = s50)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 fill = "lightgreen",
                 color = "black",
                 alpha = 0.7) +
  stat_function(fun = height_density, # Use your custom function here
                color = "darkblue",
                linewidth = 1.2,
                linetype = "solid") +
  labs(title = "Locat tree at site 50",
       x = "Height",
       y = "Density") +
  theme_minimal()
hist3 <- ggplot(height_t_df, aes(x = s80)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 fill = "lightgreen",
                 color = "black",
                 alpha = 0.7) +
  stat_function(fun = height_density, # Use your custom function here
                color = "darkblue",
                linewidth = 1.2,
                linetype = "solid") +
  labs(title = "Locat tree at site 80",
       x = "Height",
       y = "Density") +
  theme_minimal()
combined_hist <- hist1 + hist2 + hist3
combined_hist <- combined_hist +
  plot_annotation(
    title = "simbac_ARG(100, 5, 100, delta = 5, node_max = 100000)")
print(combined_hist)
```

### Number of Nodes in ARGs

Set the number of leaf lineages to be 20, 100, 200. And set $\delta = 1, 10, 20$. Increase $\rho$ from 2 to 10. For every point, I simulated 10 times and take the median value.

```{r}
#| echo: true
n_values <- c(20, 100, 200)
rho_values <- seq(2, 10, by = 2)
delta_values <- c(1, 10, 20)
func_vec <- c("sim_FSM_ARG(optimise=T)", "sim_FSM_ARG(optimise=F)", "simbac_ARG")
node_df <- expand.grid(
  n_node = NA,
  n = n_values,
  rho = rho_values,
  delta = delta_values,
  func = func_vec
)

set.seed(100)
for (i in 1:nrow(node_df)) {
  n <- node_df$n[i]
  rho <- node_df$rho[i]
  delta <- node_df$delta[i]
  func_str <- node_df$func[i]
  node_vec <- rep(NA, 10)
  if (func_str == "sim_FSM_ARG(optimise=T)") {
    for (j in 1:10) {
      r <- sim_FSM_ARG(n, rho, 100, bacteria = TRUE, delta = delta,
                       node_max = 100000, optimise_recomb = TRUE)
      node_vec[j] <- length(r$node_height)
    }
  } else if (func_str == "sim_FSM_ARG(optimise=F)") {
    for (j in 1:10) {
      r <- sim_FSM_ARG(n, rho, 100, bacteria = TRUE, delta = delta,
                       node_max = 100000, optimise_recomb = FALSE)
      node_vec[j] <- length(r$node_height)
    }
  } else if (func_str == "simbac_ARG") {
    for (j in 1:10) {
      r <- simbac_ARG(n, rho, 100, delta, node_max = 100000)
      node_vec[j] <- length(r$node_height)
    }
  }
  node_df$n_node[i] <- median(node_vec)
  print(paste("Complete", i, "iterations"))
}
```

```{r}
#| echo: true
node_plot <- node_df
node_plot$n <- paste0("n = ", node_plot$n)
node_plot$delta <- paste0("delta = ", node_plot$delta)

ggplot(node_plot, aes(x=rho, y=n_node, color=func)) +
  geom_line(aes(group = interaction(n, delta, func)), linewidth = 1.2, alpha=0.7) +
  geom_point(size = 2, shape = 21, fill = "white", stroke = 1.2, alpha=0.7) +
  scale_color_manual(values=c("sim_FSM_ARG(optimise=T)"="darkblue",
                              "sim_FSM_ARG(optimise=F)"="darkred",
                              "simbac_ARG"="darkgreen")) +
  facet_grid(n ~ delta) +
  labs(
    title = "Number of Nodes by Recombination Rates",
    x = "rho",
    y = "# of Nodes",
    color = "function"
  ) +
  scale_y_continuous(trans='log10') +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 16),
    axis.text.x = element_text(size = 12),
    axis.title.y = element_text(size = 16),
    axis.text.y = element_text(size = 12),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),
    legend.position = "bottom"
  )
```

### Running Time

Compare running time from `sim_FSM_ARG(optimise=T)`, `sim_FSM_ARG(optimise=F)`, and `simbac_ARG` by setting the number of leaf lineages = 100, number of sites = 100, $\delta = 10$, and $\rho = 10$.

```{r}
#| echo: true
benchmark_with_opt <- microbenchmark(
  sim_FSM_ARG(100, 10, 100, bacteria = TRUE, delta = 10,
              node_max = 100000, optimise_recomb = TRUE),
  times = 10,
  setup=set.seed(10)
)

benchmark_without_opt <- microbenchmark(
  sim_FSM_ARG(100, 10, 100, bacteria = TRUE, delta = 10,
              node_max = 100000, optimise_recomb = FALSE),
  times = 10,
  setup=set.seed(10)
)

benchmark_simbac <- microbenchmark(
  simbac_ARG(100, 10, 100, delta = 10, node_max = 100000),
  times = 10,
  setup=set.seed(10)
)

benchmark_with_opt
benchmark_without_opt
benchmark_simbac
```


