---
title: "Visualise ABC methods for coala model"
output: html_document
date: "2025-02-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages and functions

The files `abc_kernel.r`, `abc_basic.r` and `abc_IEnKI.r` include functions written by myself.

```{r echo=FALSE}
library(coala)
library(ggplot2)
source("abc_functions/abc_kernel.r")
source("abc_functions/abc_basic.r")
source("abc_functions/abc_IEnKI.r")
```

## Coala model

Assuming we get the following frequency spectrum from the data:

```{r sfs}
sfs <- c(112, 57, 24, 34, 16, 29, 8, 10, 15)
```

We setup a `coala` model with uniform prior for the mutation parameter. Then run the simulation 2000 times.

```{r model}
model <- coal_model(10, 50) +
  feat_mutation(par_prior("theta", runif(1, 1, 5))) +
  sumstat_sfs()
sim_data <- simulate(model, nsim = 2000, seed = 17)
```

The `create_abc_param` and `create_abc_sumstat` functions provide data for ABC methods.

```{r create_abc}
# Getting the parameters
sim_param <- create_abc_param(sim_data, model)
# Getting the summary statistics
sim_sumstat <- create_abc_sumstat(sim_data, model)
```

## ABC methods

Run basic ABC method with uniform or Gaussian kernel.

```{r basic_abc}
basic_uniform <- abc_basic(sfs, matrix(sim_param$theta), sim_sumstat, 50,
                           kernel='uniform')
basic_gaussian <- abc_basic(sfs, matrix(sim_param$theta), sim_sumstat, 50,
kernel='Gaussian', sigma=rep(1, 9))
```

Run ABC-IEnKI method by setting $\epsilon = {10^{10}, 10^{9}, \dots, 10^{-1}, 10^{-2}}$ and $\Sigma_s = I$.

```{r IEnKI_abc}
epsilon <- 10^(seq(from=10, to=-2, by=-1))
Sigma <- diag(rep(1, length(sfs)))
abc_IEnKI_list <- abc_IEnKI(sfs, matrix(sim_param$theta),
                            as.matrix(sim_sumstat), epsilon, Sigma)
pdf_values <- dnorm(abc_IEnKI_list$sumstat[, 1],
                    mean=abc_IEnKI_list$mu_T[1], sd=abc_IEnKI_list$C_T[1, 1])
```

## Vaisualsation

We compare the ABC methods by plotting their given distributions for summary statistics.

```{r plot}
abc_df <- data.frame(
  x = rep(sim_sumstat[, 1], 3),
  y = c(basic_uniform$weights, basic_gaussian$weights, pdf_values),
  type = rep(c("ABC uniform kernel", "ABC Gaussian kernel", "ABC-IEnKI"),
             each = nrow(sim_sumstat))
)

ggplot(abc_df, aes(x = x, y = y, color = type)) +
  geom_point(data = subset(abc_df, type != "ABC-IEnKI"), size = 3, alpha=0.5) +
  geom_line(data = subset(abc_df, type == "ABC-IEnKI"), linewidth = 1.2) +
  scale_color_manual(values = c("red", "blue", "green")) +
  labs(title = "ABC methods", x = "X-axis", y = "Y-axis") +
  theme_minimal()


```


