---
title: "Visualise ABC methods for coala model"
output: html_document
date: "2025-02-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages and functions

The files `abc_kernel.r`, `abc_basic.r` and `abc_IEnKI.r` include functions written by myself.

```{r load, echo=TRUE, results='hide'}
library(coala)
library(ggplot2)
library(patchwork)
source("abc_kernel.r")
source("abc_basic.r")
source("abc_IEnKI.r")
```

## Coala model

We start define a coalescent model with mutation rate $= 3$ as the target model. Assume we observe its frequency spectrum from one simulation:

```{r target}
target_model <- coal_model(10, 50) +
  feat_mutation(rate = 3) +
  sumstat_sfs()

target_sumstats <- simulate(target_model, seed=123)
sfs <- target_sumstats$sfs
print(sfs)
```

We setup a `coala` model with uniform prior for the mutation parameter. Then run the simulation 100 times.

```{r model}
model <- coal_model(10, 50) +
  feat_mutation(par_prior("theta", runif(1, 1, 5))) +
  sumstat_sfs()
sim_data <- simulate(model, nsim = 100, seed = 123)
```

The `create_abc_param` and `create_abc_sumstat` functions provide data for ABC methods.

```{r create_abc}
# Getting the parameters
sim_param <- create_abc_param(sim_data, model)
# Getting the summary statistics
sim_sumstat <- create_abc_sumstat(sim_data, model)
```

## ABC methods

Run basic ABC method with uniform or Gaussian kernel.

```{r basic_abc}
basic_uniform <- abc_basic(sfs, matrix(sim_param$theta), sim_sumstat, 50,
                           kernel='uniform')
basic_gaussian <- abc_basic(sfs, matrix(sim_param$theta), sim_sumstat, 50,
kernel='Gaussian', sigma=diag(rep(1, 9)))
```

Run ABC-IEnKI method by setting $\epsilon = {10^{10}, 10^{9}, \dots, 10^{-1}, 10^{-2}}$ and $\Sigma_s = I$.

```{r IEnKI_abc}
epsilon <- 10^(seq(from=10, to=-2, by=-1))
Sigma <- diag(rep(1, length(sfs)))
abc_IEnKI_list <- abc_IEnKI(sfs, matrix(sim_param$theta),
                            as.matrix(sim_sumstat), epsilon, Sigma)
```

## Vaisualsation

We compare the ABC methods by plotting their given distributions for summary statistics.

```{r plot, fig.width=30, fig.height=18}
plots <- lapply(1:9, function(i) {
  pdf_values <- dnorm(abc_IEnKI_list$sumstat[, i],
                      mean=abc_IEnKI_list$mu_T[i], sd=abc_IEnKI_list$C_T[i, i])
  abc_df <- data.frame(
    x = rep(sim_sumstat[, i], 3),
    y = c(basic_uniform$weights, basic_gaussian$weights, pdf_values),
    type = rep(c("ABC uniform kernel", "ABC Gaussian kernel", "ABC-IEnKI"),
                 each = nrow(sim_sumstat))
  )

  ggplot(abc_df, aes(x = x, y = y, color = type)) +
    geom_point(data=subset(abc_df, type != "ABC-IEnKI"), size=3, alpha=0.5) +
    geom_line(data=subset(abc_df, type == "ABC-IEnKI"), linewidth=1.2) +
    scale_color_manual(values=c("red", "blue", "green")) +
    labs(x=colnames(sim_sumstat)[i], y="f(s|param)") +
    theme_minimal()
})

wrap_plots(plots, ncol = 3) +
  plot_annotation(title="ABC methods",
                  theme=theme(plot.title=element_text(size=25)))
```
